---
title: "DataManipulation"
author: "fangchao"
date: "2024-10-03"
output: html_document
editor_options: 
  markdown: 
    wrap: 80
---

If this is your first time to execute this file, please pay attention to check
the availability of following packages.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("ggplot2"))
  install.packages("ggplot2")

if (!require("dplyr"))
  install.packages("dplyr")

if (!require("readr"))
  install.packages("readr")

if (!require("stringr"))
  install.packages("stringr")

```

# 1 Data collection

Before start, please read the guidance from wiki and make sure the metadata of
PRJNA636748 are already downloaded and stored under the same folder with this
file.

Try load metadata into R

```{r load}
metadata.tb <- read_csv("SraRunTable.txt")
```

# 2 Data landscapge

Summary:

```{r summary}
summary(metadata.tb)
```

## 2.1 Geography distribution

If we are curious about the samples collected from different regions:

Using `table`:

```{r}
table(metadata.tb$geo_loc_name)
```

sorted:

```{r}
sort(table(metadata.tb$geo_loc_name))
```

Try uniform the names for better grouping:

```{r}

# Old fashion way
metadata.tb$geo_loc_name[which(metadata.tb$geo_loc_name=="South Africa: unknown")] <- "South Africa: Unknown"

# dplyr style
metadata.fix.df <- metadata.tb %>%
  mutate(geo_loc_name = str_replace_all(geo_loc_name, 
    c("South Africa: Kwazulu-Natal" = "South Africa: KwaZulu-Natal", 
      "South Africa: KwaZulu Natal" = "South Africa: KwaZulu-Natal",
      "South Africa: Kwazulu natal" = "South Africa: KwaZulu-Natal",
      "South Africa: KZN"           = "South Africa: KwaZulu-Natal",
      "South Africa: nan"           = "South Africa: Unknown",
      "South Africa: missing"       = "South Africa: Unknown",
      "South Africa: Freestate"     = "South Africa: Free State",
      "missing" = "Unknown")))

sort(table(metadata.fix.df$geo_loc_name))
```

To visualize sample counts among geograpic locations

```{r}
p <- ggplot(metadata.fix.df,aes(x=geo_loc_name)) + geom_bar()

p
```

Add `coord_flip` so the legend can be displayed better

```{r}
p + coord_flip()
```

How to reorder?

```{r}
revorder <- sort(table(metadata.fix.df$geo_loc_name))
revorder
metadata.fix.df$geo_loc_name <- factor(metadata.fix.df$geo_loc_name,levels=names(revorder))
```

```{r}
ggplot(metadata.fix.df,aes(x=geo_loc_name)) + geom_bar() + coord_flip()
```

## 2.2 Platform / Instrument distribution

```{r}
table(metadata.fix.df$Instrument,metadata.fix.df$Platform)
```

Visualization:

```{r}

ins_order <- sort(table(metadata.fix.df$Instrument))
metadata.fix.df$Instrument <- factor(metadata.fix.df$Instrument,levels=names(ins_order))
ggplot(metadata.fix.df,aes(x=Instrument)) +
  geom_bar(aes(fill=Platform))
```

## 2.3 Sequencing volume distribution

```{r}

ggplot(metadata.fix.df,aes(x=Bytes)) + geom_density()
```

Characteristics - Skewed Distribution - Two peaks, indicating two main
distributions mixed;

Solution: - log-transform - Find out the variable where the two distributions
came from.

```{r}
ggplot(metadata.fix.df,aes(x=Bytes)) + geom_density() + 
  scale_x_log10() + ggtitle("log10 scaled distribution")
  

ggplot(metadata.fix.df,aes(x=Bytes,color=Instrument)) + geom_density() +
  ggtitle("Instrument colored distribution")


ggplot(metadata.fix.df,aes(x=Bytes,color=Instrument)) + geom_density() + xlim(c(0,2e8)) +
  ggtitle("Instrument colored distribution")

```

```{r}
table(metadata.fix.df$Instrument,metadata.fix.df$LibraryLayout)
```

## 2.4 Collectors

```{r}
sort(names(table(metadata.fix.df$collected_by)))
```

Naming uniform: CDC - Africa Centers for Disease Control and Prevention AHRI -
Africa Health Research Institute AMPATH - Academic Model Providing Access to
Healthcare BARC - Bio Analytical Research Corporation CAPRISA - Centre for the
AIDS Programme of Research in South Africa MRC - Medical Research Council NHLS -
National Health Laboratory Service Netcare -

```{r}
metadata.fix.df$collected_by <- as.factor(metadata.fix.df$collected_by)
```

Check the factor's levels:

```{r}
levels(metadata.fix.df$collected_by)
```

rename the levels:

```{r}
levels(metadata.fix.df$collected_by) <- c(
     "CDC SA",
    "AHRI",
    "AHRI",
    "AHRI",
    "AHRI",
    "Ampath",
    "Ampath",
    "Ampath",
    "AMPATH",
    "BARC",
    "BARC",
    "BARC",
    "CAPRISA",
    "CAPRISA",
    "CAPRISA",
    "CAPRISA",
    "Stellenbosch University",
    "LANCET",
    "missing",
    "Molecular Diagnostics Services",
    "MRC",
    "NHLS",
    "NHLS",
    "NHLS",
    "NHLS",
    "NHLS",
    "NHLS",
    "NHLS",
    "Netcare",
    "Netcare",
    "Netcare",
    "Netcare",
    "Netcare",
    "Netcare",
    "Netcare",
    "NHLS",
    "NHLS",
    "NHLS",
    "NHLS",
    "NHLS",
    "PathCare",
    "PathCare",
    "University of Witwatersand",
    "University of Witwatersand",
    "University of Witwatersand",
    "BARC",
    "Stellenbosch University",
    "Univeristy of Pretoria",
    "University of Cape-Town",
    "University of Freestate",
    "University of Pretoria",
    "University of Witwatersand"
)
```

```{r}
levels(metadata.fix.df$collected_by)
```

```{r}
revorder <- sort(table(metadata.fix.df$collected_by))
revorder
metadata.fix.df$collected_by <- factor(metadata.fix.df$collected_by,levels=names(revorder))


ggplot(metadata.fix.df,aes(x=collected_by)) + geom_bar() + coord_flip()
  ggtitle("Instrument colored distribution")
```

boxplot

```{r}
ggplot(metadata.fix.df,aes(x=collected_by,y=Bytes)) +
  geom_boxplot() + coord_flip()
```

```{r}
ggplot(metadata.fix.df,aes(x=collected_by,y=Bytes,fill=Instrument)) +
  geom_boxplot() + coord_flip() + ylim(c(0,1e7))
```

# 3 Subsampling dataset

Fell free to modify this section to fit your interested question:)

Example question: Does the virus strains different among geographic regions? -
Number of samples picking for each group: 3 - Only use Illumina MiSeq to avoid
bias between instruments

```{r}
df_subsampled <- metadata.fix.df %>%
  filter(Instrument=="Illumina MiSeq" & 
           collected_by %in% c("NHLS","PathCare")) %>%
  group_by(geo_loc_name) %>%
  sample_n(size = 3) %>%
  ungroup()
```

`filter(Instrument=="Illumina MiSeq")`: Filters the data to keep only the rows
where the Instrument column is equal to "Illumina MiSeq".

`group_by(geo_loc_name)`: Groups the data by the `geo_loc_name column`.

`sample_n(size = 3)`: Randomly samples 3 rows from each group created by the
previous group_by() operation.

`ungroup()`: Removes the grouping structure from the data frame. This is often
done after a grouped operation to ensure that subsequent operations are
performed on the entire data frame and not on grouped subsets.

```{r}
ggplot(df_subsampled,aes(x=geo_loc_name,y=Bytes)) + 
  geom_boxplot() + geom_jitter(width=.2) + coord_flip()
```

Export:
```{r}
write.csv(df_subsampled,"subsample_metadata.csv")
```

This is all for today!  
In the next course, we will attempt to download above subsampled dataset and execute a basic pipeline.  
